Using Linear Regression to Examine the Correlation Between Access to Drinking Water and Life Expectancy

### 2022-12-09 Manuscript/Paper.tex
add figures
clean repo
change figures file type
change width of figures
additions

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage[margin=1 in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{lipsum}
\usepackage[colorlinks=true, citecolor=blue]{hyperref}
\usepackage{listings}
\usepackage{changepage}
\setlength{\parindent}{0pt}

\title{Using Linear Regression to Examine the Correlation Between Access to Drinking Water and Life Expectancy}
\author{Julia Andronowitz\\
University of Connecticut\\Statistics Department}
\date{November 14, 2022}

@@

\maketitle

@@

@@

Living in the United States, we often take the necessity of safe drinking water for granted, but it is still a pressing issue in many parts of the world. Current research shows that unsafe water is responsible for millions of deaths \citep{ritchieroser2019water}. From the \textit{World Health Organization}, "Contaminated water and poor sanitation are linked to transmission of diseases such as cholera, diarrhoea, dysentery, hepatitis A, typhoid and polio" \citep{who2017water}. These diseases are largely preventable, and an increase in both access to and quality of safe drinking water can dramatically increase life expectancy \citep{angelakis2021quality}. Such a study could help emphasize the need for clean water worldwide, as it has numerous health benefits beyond longevity \citep{popkin2010waterhealth}. \\

@@

The two data sets we will be using are from the World Health Organization and provided by Kaggle. The first is called Basic Drinking Water Services and contains information regarding basic drinking water services across various countries and in different years. There are 3455 rows with 4 columns: Location, Period, Indicator, and Tooltip. The indicator column is a text cell that says "Population using at least basic drinking-water services", and the Tooltip column gives the percentage of the population where the indicator is true. In this data set, we will be looking at the country, year, and percentage of population using at least basic drinking water services. \\

@@

\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
import numpy as np
import pandas as pd
import seaborn as sns
from bokeh.plotting import figure
from bokeh.io import show, output_notebook
from bokeh.models import ColumnDataSource, HoverTool
from bokeh.layouts import gridplot
output_notebook()
\end{lstlisting}
\end{adjustwidth}

Next, we import the data:
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
rawData1=pd.read_csv('archive/basicDrinkingWaterServices.csv')
rawData2=pd.read_csv('archive/lifeExpectancyAtBirth.csv')
\end{lstlisting}
\end{adjustwidth}

If we would like to see the data, use the command
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
rawData1
\end{lstlisting}
\end{adjustwidth}

An output of the data is below:

\begin{center}
\includegraphics[width=5in]{figure1.pdf}
\includegraphics[width=5in]{figure2.pdf}
\end{center}

For the drinking water data, we take only the columns with the country, year, and the percentage of the population using at least basic drinking water services.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
rawData1=rawData1[["Location", "Period" , "First Tooltip"]]
\end{lstlisting}
\end{adjustwidth}

We do the same for the life expectancy data, only taking the columns with the country, year, sex, and average life expectancy at birth.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
rawData2=rawData2[["Location", "Period", "Dim1", "First Tooltip"]]
\end{lstlisting}
\end{adjustwidth}

Next, we want to convert our pandas dataframes into numpy arrays so that we can use the appropriate mathematical functions.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
data1=np.array(rawData1)
data2=np.array(rawData2)
\end{lstlisting}
\end{adjustwidth}

We want to have as many constants as possible in our model. Since the data was collected starting in 2000, we must use the same years in both data sets. We will use the year 2015 as the benchmark. The data can then be filtered using a loop and appending the 2015 data to a new list.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
dw2015=[]
for x in data1:
    if x[1]==2015:
        dw2015.append(x)
\end{lstlisting}
\end{adjustwidth}

This must be done for both data sets. We now do it for the life expectancy data set. As previously mentioned in the "Data" section, we also want to use only the rows with both sexes.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
le2015=[]
for x in data2:
    if (x[1]==2015) and (x[2]=='Both sexes'):
        le2015.append(x)
\end{lstlisting}
\end{adjustwidth}

These lists will also be converted into numpy arrays.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
DW2015=np.array(dw2015)
lifeExpectancy2015=np.array(le2015)
\end{lstlisting}
\end{adjustwidth}

Now, we see the shapes of both these arrays are different. The drinking water array contains 193 rows and 3 columns while the life expectancy array contains 183 rows and 4 columns. This means that there are some countries that are not in both lists. So, we need to delete the data for countries that are only in one or the other. We do this by checking if the countries in the drinking water array are also in the life expectancy data. If both countries are in both data sets, we append them to our new drinking water list, which is what we will use going forward.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
drinkingWater2015=[]
for x in DW2015:
    if x[0] in lifeExpectancy2015:
        drinkingWater2015.append(x)
drinkingWater2015=np.array(drinkingWater2015)
\end{lstlisting}
\end{adjustwidth}

Now, we can see the lists are the same length. If the lengths were still different, we could do the same process with the life expectancy list, i.e. making sure every country in the life expectancy list is also in the drinking water list and removing elements that are not. Next, we want to assign $x$ to the percentages of people with access to basic drinking water services and $y$ to the life expectancies in 2015.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
x2015=drinkingWater2015[:,2]
x2015=x2015.astype('float64')
y2015=lifeExpectancy2015[:,3]
y2015=y2015.astype('float64')
\end{lstlisting}
\end{adjustwidth}

Now that we have standardized data across the board, we can plot the life expectancy and basic drinking water.

\vspace{-130pt}

\begin{center}
\includegraphics[width=6in]{figure3.pdf}\\
\end{center}

\vspace{-160pt}

\begin{center}
    \textit{Figure 1.}
\end{center}

\vspace{20pt}

Now we create our data matrix $x$, which is the $x$ array transformed into a column vector and appended to a column of ones.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
x2015=x2015.reshape(-1,1)
X2015=np.concatenate([x2015,np.ones(shape=(x2015.shape[0],1))],axis=1)
\end{lstlisting}
\end{adjustwidth}

Since this is now a column vector, we also want to transform $y$ into a column vector.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
Y2015=y2015.reshape(-1,1)
\end{lstlisting}
\end{adjustwidth}

Now, we can compute the least squares line and the matrix $M$ that will give us the slope and the $y$-intercept for the line of best fit.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
D2015=X2015.transpose()@X2015
np.linalg.inv(D2015)
M2015=(np.linalg.inv(D2015)@X2015.transpose())@Y2015
\end{lstlisting}
\end{adjustwidth}

The output is the array $[[0.27859093],[49.15736534]]$, which tells us the slope of the line is 0.279 and the intercept is 49.157. If we want to compare the actual values to predicted values, we use the following code:
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
Yhat2015 = np.dot(np.dot(np.dot(X2015,np.linalg.inv(D2015)),
                                    X2015.transpose()),Y2015)
\end{lstlisting}
\end{adjustwidth}

A visual representation of the least squares line can be seen once executing
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
f2015.line(x=x2015.ravel(),y=Yhat2015[:,0],color='black')
show(f2015)
\end{lstlisting}
\end{adjustwidth}

\vspace{-130pt}

\begin{center}
\includegraphics[width=6in]{figure4.pdf}
\end{center}

\vspace{-160pt}

\begin{center}
    \textit{Figure 2.}
\end{center}

\vspace{20pt}

@@

As seen in \textit{Figure 1}, we can assume a linear regression is appropriate for the model based on the data. There looks to be a linear correlation between the variables rather than a quadratic or other polynomial relationship. To see the R-Squared value of our linear regression, we can use the following code \citep{pythonr2}:
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
corr_matrix = np.corrcoef(Y2015[:,0], Yhat2015[:,0])
corr = corr_matrix[0,1]
R_sq = corr**2
\end{lstlisting}
\end{adjustwidth}
We see the $R^2$ value is 0.66348. So, 66.348\% of the variability in the data is explained by the model. While this is on the lower end, our model is still supported. We would also like to check that the mean of the residuals is close to zero. This can be calculated with the following code.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
residuals=Yhat2015[:,0]-Y2015[:,0]
np.mean(residuals)
\end{lstlisting}
\end{adjustwidth}
We see the mean is -1.2813065726821479e-14 which is very close to zero. We also need to check for homosteadasticity among the residuals. To look at the residuals, the difference between the measured and predicted values, we can use the following function as detailed in \citet{teitelbaum2021linreg}.
\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}[language=Python]
def comparison_plot(x,Y, Yhat):
    '''Plots Predicted vs True values for analysis of regression'''
    comparison_plot = 
        figure(title='Difference between Measured and Predicted Values')
    comparison_plot.xaxis.axis_label='x'
    comparison_plot.yaxis.axis_label='Yhat-Y'
    comparison_plot.scatter(x=x,y=Yhat-Y)
    comparison_plot.line(x=[x.min(),x.max()],y=[0,0])
    return comparison_plot
show(comparison_plot(x2015.ravel(),Y2015[:,0], Yhat2015[:,0]))
\end{lstlisting}
\end{adjustwidth}

\vspace{-40pt}

\begin{center}
\includegraphics[width=4in]{figure5.pdf}
\end{center}

\vspace{-50pt}

\begin{center}
    \textit{Figure 3.}
\end{center}

We see there are no patterns among the residuals, so this assumption is validated.

\vspace{15pt}

@@

\begin{adjustwidth}{0.5in}{0.5in}
\begin{lstlisting}
p = sns.displot(residuals,kde=True)
\end{lstlisting}
\end{adjustwidth}

\vspace{-150pt}

\begin{center}
    \includegraphics[width=6in]{figure6.pdf}
\end{center}

\vspace{-160pt}

\begin{center}
\textit{Figure 4.}
\end{center}

\vspace{15pt}

We see the graph is slightly bimodal but almost normally distributed as it is very hard to get perfectly normally distributed curves with real life.

\vspace{15pt}

@@

\begin{adjustwidth}{0.5in}{0.25in}
\begin{lstlisting}
df = pd.DataFrame({'x': x2015[:,0],
                   'y': Y2015[:,0]})
model = ols('x ~ y', data=df).fit()
stud_res = model.outlier_test()

for x in stud_res['student_resid']:
    if x > 3 or x < -3:
        print(x)
\end{lstlisting}
\end{adjustwidth}

@@

\vspace{15pt}

@@

\bibliography{references}
\bibliographystyle{chicago}

>>>>

\documentclass[12pt]{article}

\usepackage{amsmath}
\usepackage[margin=1 in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage[colorlinks=true, citecolor=blue]{hyperref}
\usepackage{listings}
\usepackage{minted}
\usepackage{hyperref}

\parindent=0pt
\parskip=\bigskipamount

\title{Using Linear Regression to Examine the Correlation Between Access to Drinking Water and Life Expectancy}
\author{Julia Andronowitz\\B.S., Mathematics and Statistics}
\date{December 9, 2022}

@@

\maketitle
\thispagestyle{empty}

@@

\vfill

\begin{center}
    University of Connecticut \\ Department of Statistics \\ Professor Jun Yan, Ph.D.
\end{center}

@@

Living in the United States, we often take the necessity of safe drinking water for granted, but it is still a pressing issue in many parts of the world. Current research shows that unsafe water is responsible for millions of deaths \citep{ritchieroser2019water}. From the \textit{World Health Organization}, "Contaminated water and poor sanitation are linked to transmission of diseases such as cholera, diarrhoea, dysentery, hepatitis A, typhoid and polio" \citep{who2017water}. These diseases are largely preventable, and an increase in both access to and quality of safe drinking water can dramatically increase life expectancy \citep{angelakis2021quality}.

Although there are numerous studies on how drinking contaminated water can have tremendous negative effects on one's health, there are very few studies that examine how the percentage of people with access to clean water in a certain area affects statistics regarding the community's overall health. Previous studies regarding drinking water and health often use the presence of clean water as a binary outcome, where a community either does or does not have access. Other times, facilities lie on a continuum. In \citet{wolf2014sanitationondisease}, the quality of drinking water was measured by water and sanitation facilities that lie on a scale between poor and good. Conclusions were drawn based on the improvements to water systems, stating that "...there are large potential reductions in diarrhoeal disease risk through improvements to both water and sanitation in low- and middle-income settings" \citep{wolf2014sanitationondisease}.

Additionally, studies regarding drinking water have been done to draw conclusions about the presence of certain diseases when clean drinking water is limited. Yet, there are other external effects of maintaining access to clean drinking water. From \citet{2020mappinggeo}, "Low access to safe water and sanitation has also been linked to broader social outcomes such as reductions in school attendance (particularly for girls who are menstruating), losses to economic productivity, and undue burden on women of time spent collecting water". This study aims to quantify how access to drinking water affects a population's overall life expectancy at birth. Such a study could help emphasize the need for clean water worldwide, as it has numerous health benefits beyond longevity \citep{popkin2010waterhealth}.

@@

The two data sets we will be using are from the World Health Organization and provided by Kaggle. The first is called Basic Drinking Water Services and contains information regarding basic drinking water services across various countries and in different years. There are 3455 rows with 4 columns: Location, Period, Indicator, and Tooltip. The indicator column is a text cell that says "Population using at least basic drinking-water services", and the Tooltip column gives the percentage of the population where the indicator is true. In this data set, we will be looking at the country, year, and percentage of population using at least basic drinking water services.

@@

\begin{minted}{python}
    import numpy as np
    import pandas as pd
    import seaborn as sns
    from bokeh.plotting import figure
    from bokeh.io import show, output_notebook
    from bokeh.models import ColumnDataSource, HoverTool
    from bokeh.layouts import gridplot
    output_notebook()
\end{minted}

Next, we import the data:
\begin{minted}{python}
    rawData1=pd.read_csv('archive/basicDrinkingWaterServices.csv')
    rawData2=pd.read_csv('archive/lifeExpectancyAtBirth.csv')
\end{minted}

If we would like to see the data, use the command
\begin{minted}{python}
    rawData1
\end{minted}

An output of the data is in Figures ~\ref{fig1} and ~\ref{fig2}.
\begin{figure}
    \centering
    \includegraphics[width=5in]{Figures/data1.pdf}
    \caption{Access to Basic Drinking Water Services Data}
    \label{fig1}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=5in]{Figures/data2.pdf}
    \caption{Life Expectancy at Birth Data}
    \label{fig2}
\end{figure}
For the drinking water data, we take only the columns with the country, year, and the percentage of the population using at least basic drinking water services.
\begin{minted}{python}
    rawData1=rawData1[["Location", "Period" , "First Tooltip"]]
\end{minted}

We do the same for the life expectancy data, only taking the columns with the country, year, sex, and average life expectancy at birth.
\begin{minted}{python}
    rawData2=rawData2[["Location", "Period", "Dim1", "First Tooltip"]]
\end{minted}

Next, we want to convert our pandas dataframes into numpy arrays so that we can use the appropriate mathematical functions.
\begin{minted}{python}
    data1=np.array(rawData1)
    data2=np.array(rawData2)
\end{minted}

We want to have as many constants as possible in our model. Since the data was collected starting in 2000, we must use the same years in both data sets. We will use the year 2015 as the benchmark. The data can then be filtered using a loop and appending the 2015 data to a new list.
\begin{minted}{python}
    dw2015=[]
    for x in data1:
        if x[1]==2015:
            dw2015.append(x)
\end{minted}

This must be done for both data sets. We now do it for the life expectancy data set. As previously mentioned in the "Data" section, we also want to use only the rows with both sexes.
\begin{minted}{python}
    le2015=[]
    for x in data2:
        if (x[1]==2015) and (x[2]=='Both sexes'):
            le2015.append(x)
\end{minted}

These lists will also be converted into numpy arrays.
\begin{minted}{python}
    DW2015=np.array(dw2015)
    lifeExpectancy2015=np.array(le2015)
\end{minted}

Now, we see the shapes of both these arrays are different. The drinking water array contains 193 rows and 3 columns while the life expectancy array contains 183 rows and 4 columns. This means that there are some countries that are not in both lists. So, we need to delete the data for countries that are only in one or the other. We do this by checking if the countries in the drinking water array are also in the life expectancy data. If both countries are in both data sets, we append them to our new drinking water list, which is what we will use going forward.
\begin{minted}{python}
    drinkingWater2015=[]
    for x in DW2015:
        if x[0] in lifeExpectancy2015:
            drinkingWater2015.append(x)
    drinkingWater2015=np.array(drinkingWater2015)
\end{minted}

Now, we can see the lists are the same length. If the lengths were still different, we could do the same process with the life expectancy list, i.e. making sure every country in the life expectancy list is also in the drinking water list and removing elements that are not. Next, we want to assign $x$ to the percentages of people with access to basic drinking water services and $y$ to the life expectancies in 2015.
\begin{minted}{python}
    x2015=drinkingWater2015[:,2]
    x2015=x2015.astype('float64')
    y2015=lifeExpectancy2015[:,3]
    y2015=y2015.astype('float64')
\end{minted}

Now that we have standardized data across the board, we can plot the life expectancy and basic drinking water. The output is shown in Figure ~\ref{fig3}.
\begin{figure}
    \centering
    \includegraphics[width=4in]{Figures/figure1.png}
    \caption{Life Expectancy vs. Drinking Water}
    \label{fig3}
\end{figure}

Now we create our data matrix $x$, which is the $x$ array transformed into a column vector and appended to a column of ones.
\begin{minted}{python}
    x2015=x2015.reshape(-1,1)
    X2015=np.concatenate([x2015,np.ones(shape=(x2015.shape[0],1))],axis=1)
\end{minted}

Since this is now a column vector, we also want to transform $y$ into a column vector.
\begin{minted}{python}
    Y2015=y2015.reshape(-1,1)
\end{minted}

Now, we can compute the least squares line and the matrix $M$ that will give us the slope and the $y$-intercept for the line of best fit.
\begin{minted}{python}
    D2015=X2015.transpose()@X2015
    np.linalg.inv(D2015)
    M2015=(np.linalg.inv(D2015)@X2015.transpose())@Y2015
\end{minted}

The output is the array $[[0.27859093],[49.15736534]]$, which tells us the slope of the line is 0.279 and the intercept is 49.157. If we want to compare the actual values to predicted values, we use the following code:
\begin{minted}{python}
    Yhat2015 = np.dot(np.dot(np.dot(X2015,np.linalg.inv(D2015)),
                                    X2015.transpose()),Y2015)
\end{minted}

A visual representation of the least squares line can be seen once executing
\begin{minted}{python}
    f2015.line(x=x2015.ravel(),y=Yhat2015[:,0],color='black')
    show(f2015)
\end{minted}
The output is shown in Figure ~\ref{fig4}
\begin{figure}
    \centering
    \includegraphics[width=4in]{Figures/figure2.png}
    \caption{Plot with Least Squares Line}
    \label{fig4}
\end{figure}

@@

As seen in Figure ~\ref{fig3}, we can assume a linear regression is appropriate for the model based on the data. There looks to be a linear correlation between the variables rather than a quadratic or other polynomial relationship. To see the R-Squared value of our linear regression, we can use the following code \citep{pythonr2}:
\begin{minted}{python}
    corr_matrix = np.corrcoef(Y2015[:,0], Yhat2015[:,0])
    corr = corr_matrix[0,1]
    R_sq = corr**2
\end{minted}
We see the $R^2$ value is 0.66348. So, 66.348\% of the variability in the data is explained by the model. While this is on the lower end, our model is still supported. We would also like to check that the mean of the residuals is close to zero. This can be calculated with the following code.
\begin{minted}{python}
    residuals=Yhat2015[:,0]-Y2015[:,0]
    np.mean(residuals)
\end{minted}
We see the mean is -1.2813065726821479e-14 which is very close to zero. We also need to check for homosteadasticity among the residuals. To look at the residuals, the difference between the measured and predicted values, we can use the following function as detailed in \citet{teitelbaum2021linreg}.
\begin{minted}{python}
    def comparison_plot(x,Y, Yhat):
        '''Plots Predicted vs True values for analysis of regression'''
        comparison_plot = 
            figure(title='Difference between Measured and Predicted Values')
        comparison_plot.xaxis.axis_label='x'
        comparison_plot.yaxis.axis_label='Yhat-Y'
        comparison_plot.scatter(x=x,y=Yhat-Y)
        comparison_plot.line(x=[x.min(),x.max()],y=[0,0])
        return comparison_plot
    show(comparison_plot(x2015.ravel(),Y2015[:,0], Yhat2015[:,0]))
\end{minted}

\begin{figure}
    \centering
    \includegraphics[width=4in]{Figures/figure3.png}
    \caption{Residuals Plot}
    \label{fig5}
\end{figure}

In Figure ~\ref{fig5}, we see there are no patterns among the residuals, so this assumption is validated.

@@

\begin{minted}{python}
    p = sns.displot(residuals,kde=True)
\end{minted}

\begin{figure}
    \centering
    \includegraphics[width=4in]{Figures/figure4.png}
    \caption{Distribution Plot}
    \label{fig6}
\end{figure}

We see the graph in Figure ~\ref{fig6} is slightly bimodal but almost normally distributed as it is very hard to get perfectly normally distributed curves with real life.

@@

\begin{minted}{python}
    df = pd.DataFrame({'x': x2015[:,0],
                       'y': Y2015[:,0]})
    model = ols('x ~ y', data=df).fit()
    stud_res = model.outlier_test()

    for x in stud_res['student_resid']:
        if x > 3 or x < -3:
            print(x)
\end{minted}

@@

@@

\bibliography{References/References.bib}
\bibliographystyle{References/chicago.bst}
